Large squeezing in multicomponent trapped Bose-Einstein condensates

### 2013-05-24 large_number_squeezing_with_atoms.tex
Minor edits.

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the first nonlinear term in the multimode Hamiltonian becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in one mode, we can approximate this sum as the single term $\hat{H} = \chi_{aa} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}/2$ where
\begin{equation}
\chi_{aa} = U_{aa} \int |u_i|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity. In the multimode case, the effective nonlinearity of a mode is a function of the mode shape, as is shown by Eq.~(\ref{eqChiUequivalence}). The optimum hold time $\tau_{\mathrm{hold}}$ depends on the product of the effective mode nonlinearity and number number of particles in that mode, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying effective nonlinearity-particle number products and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$. This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across all the spacial modes present, result in an overall lowest possible total number variance.

@@

One way to avoid any requirement to engineer and maintain specific atomic modes is to post-select a spatial mode.  Rather than considering the variance in the total number of atoms in a mode, it may be possible to consider only the statistics of a spatially filtered area of the atomic cloud.  For example, suppose the atoms were in a Gaussian atomic distribution, but we consider a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard Gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\chi = U \frac{{\mathrm{Erf}} (\sqrt{2p})^3}{{\mathrm{Erf}}(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}
\end{equation}
Spatial filtering can easily produce higher squeezing, but there is the obvious disadvantage that it  significantly reduces the effective number of atoms.  This is a serious issue for most applications of squeezed sources, such as interferometry, where the signal to noise scales with the square root of the flux.  We will focus on methods for producing squeezing with large atomic number.

@@

The Bogoliubov theory used in deriving \eqref{eqBogoliubovSqueezing} is not number-conserving because the depletion of the condensate caused by the occupation of the non-zero momentum modes is neglected.  Therefore when these non-zero momentum modes become sufficiently occupied for $S_a$ to differ from the two-mode model, the Bogoliubov approximation becomes invalid and $S_a$ will not agree with the full quantum-statistical multimode result.  While number-conserving variants of the Bogoliubov theory exist \cite{NumberConservingBog}, they are still perturbative approximations about the mean field, and they are not required for our purposes.  Although the Bogoliubov approximation cannot predict the exact amount by which squeezing will be degraded compared to the two-mode model, it does predict the regime in which squeezing is reduced by coupling between spatial modes, it can thereby suggest how this problem can be avoided.  Figure~\ref{figBogSqueezingValidation}(a) illustrates the agreement between truncated Wigner simulations and the Bogoliubov theory for the total number of atoms in the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ for parameters where these modes approach 20\% of the total number of atoms in the system.  Figure~\ref{figBogSqueezingValidation}(b) shows that in this case the squeezing predicted by the Bogoliubov theory in Eq.~\eqref{eqBogoliubovSqueezing} is a lower bound for the squeezing predicted by the truncated Wigner simulations.  The two-mode analytic solution, which ignores multimode effects entirely, provides an upper bound for the squeezing. \textbf{Reconsider these last sentences when we have the figure.  Maybe qualify `lower bound' for constant density models only?}  This means that we can use the Bogoliubov analysis to guide our search for parameter regimes with higher squeezing.

To minimise the damaging effects of the non-zero momentum modes, it is better to operate in a regime where the total occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ is small, which is also fortuitously the regime in which the Bogoliubov theory is a good approximation.  
The number of excited modes is almost entirely determined by the density of modes energetically accessible by the system.  While this gets larger with the dimensionality and the volume of the system at a given chemical potential, in practice the dimensionality and the volume of the BEC are controlled by changing the trap parameters, which has a dramatic effect on the attainable chemical potential.  So the true figure of merit controlling the degradation of the squeezing due to coupling between the modes is the number of the number of states energetically reachable from the chemical potential.  This scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the number of accessible states increases with volume, whereas for 2D traps it is constant with volume.

@@

In Fig.~\ref{fig:3DBogScaling} we see that the number of accessible states actually decreases with volume.  This is because the energy of the ground state of the BEC is reducing faster than the density of states at low energies is increasing.  \textbf{Create these figures (by combining old figures) and check both for latest updates.  Confirm that 'simulation' (blue line) means TWA, as opposed to Graham's prediction.}  The non-smooth nature of the Bogoliubov calculation is due to the coarse grid, which leads to granularity as the scale factor is changed.

>>>>

We begin by considering a simplified version of the problem, assuming that only one of these spatial modes is relevant for each of the condensate's components.  This is done by writing the multimode atomic field operators in terms of a set of basis functions $\hat{\psi}_a({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{a}_{n}(t)$ and $\hat{\psi}_b({\mathbf{r}},t) = \sum_n u_{n}({\mathbf{r}}) \hat{b}_{n}(t)$, where $\hat{a}_{n}$ and $\hat{b}_{n}$ annihilate a particle with normalised spatial mode $u_n$ and internal state $|a\rangle$ or $|b\rangle$ respectively. With this expansion, the nonlinear term in the multimode Hamiltonian for the self-interaction of the $|a\rangle$ internal state becomes
\begin{equation}
\hat{H} = \frac{U_{aa}}{2} \int dV \sum_{nmpq} u_n^* u_m^* u_p u_q \, \hat{a}^{\dagger}_{n} \hat{a}^{\dagger}_{m} \hat{a}_{p} \hat{a}_{q} 
\end{equation}
When the system remains predominantly in one mode, we can approximate this sum as the single term $\hat{H} = \chi_{aa} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}/2$ where
\begin{equation}
\chi_{aa} = U_{aa} \int |u_i|^4 \, dV.
\label{eqChiUequivalence}
\end{equation}
We proceed similarly for the other terms.  In the parameter regimes we are interested in, the detuning between the two spatial modes will be irrelevant, so setting our zero of energy appropriately, we are left with only a standard Kerr-type nonlinearity arising from atom-atom interactions, as well as a linear coupling between the two modes.  The Hamiltonian governing this reduced system is given by
\begin{equation}
\hat{H} = \frac{\chi_{aa}}{2} \hat{a}^{\dagger} \hat{a}^{\dagger} \hat{a} \hat{a}
          + \chi_{ab} \hat{a}^{\dagger} \hat{a} \hat{b}^{\dagger} \hat{b}
          + \frac{\chi_{bb}}{2} \hat{b}^{\dagger} \hat{b}^{\dagger} \hat{b} \hat{b}
          + \Omega (\hat{a}^{\dagger} \hat{b} + \hat{b}^{\dagger}  \hat{a} )
\label{eqTwoModeHamiltonian}
\end{equation}
where the $\chi_{ij}$ describe the various inter- and intra-mode nonlinearities, and $\Omega$ couples the two modes and allows for population transfer between them.

@@

The first problem is present even if the spatial modes are uncoupled.  The two-mode model shows that the parameters required to obtain best squeezing are dependent on the strength of the nonlinearity. In the multimode case, the effective nonlinearity of a mode is a function of the mode shape, as is shown by Eq.~(\ref{eqChiUequivalence}). The optimum hold time $\tau_{\mathrm{hold}}$ depends on the product of the effective mode nonlinearity and number number of particles in that mode, as does the recombination phase $\phi$.  Different modes in a multimode environment will typically have varying effective nonlinearity-particle number products and therefore squeeze at different rates, with each mode having a different optimum $\tau_{\mathrm{hold}}$ and optimum recombination phase $\phi$. This means that the best squeezing will be achieved by choosing values of $\tau_{\mathrm{hold}}$ and $\phi$ that, averaged across all the spatial modes present, result in an overall lowest possible total number variance.

@@

One way to avoid any requirement to engineer and maintain specific atomic modes is to post-select a spatial mode.  Rather than considering the variance in the total number of atoms in each internal state, it may be possible to consider only the statistics in a spatially filtered area of the atomic cloud.  For example, suppose the atoms were in a Gaussian atomic distribution, but we consider a mode that only includes the atoms within $p$ standard deviations of the centre. For low $p$, this mode shape is considerably closer to the constant density case than a standard Gaussian. The equivalence between the single mode case and this mode is given by
\begin{equation}
\chi = U \frac{{\mathrm{Erf}} (\sqrt{2p})^3}{{\mathrm{Erf}}(p)^6} \sqrt{\frac{m^3 \omega_x \omega_y \omega_z}{8 \pi^3 \hbar^3}}
\end{equation}
Spatial filtering can easily produce higher squeezing, but there is the obvious disadvantage that it  significantly reduces the effective number of atoms.  This is a serious issue for most applications of squeezed sources, such as interferometry, where the signal to noise scales with the square root of the flux.  We will focus on methods for producing squeezing with large atomic number.

@@

The Bogoliubov theory used in deriving \eqref{eqBogoliubovSqueezing} is not number-conserving because the depletion of the condensate caused by the occupation of the non-zero momentum modes is neglected.  Therefore when these non-zero momentum modes become sufficiently occupied for $S_a$ to differ from the two-mode model, the Bogoliubov approximation becomes invalid and $S_a$ will not agree with the full quantum-statistical multimode result.  While number-conserving variants of the Bogoliubov theory exist \cite{NumberConservingBog}, they are still perturbative approximations about the mean field, and they are not required for our purposes.  Although the Bogoliubov approximation cannot predict the exact amount by which squeezing will be degraded compared to the two-mode model, it does predict the regime in which squeezing is reduced by coupling between spatial modes, and it can thereby suggest how this problem can be avoided.  Figure~\ref{figBogSqueezingValidation}(a) illustrates the agreement between truncated Wigner simulations and the Bogoliubov theory for the total number of atoms in the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ for parameters where these modes approach 20\% of the total number of atoms in the system.  Figure~\ref{figBogSqueezingValidation}(b) shows that in this case the squeezing predicted by the Bogoliubov theory in Eq.~\eqref{eqBogoliubovSqueezing} is a lower bound for the squeezing predicted by the truncated Wigner simulations.  The two-mode analytic solution, which ignores multimode effects entirely, provides an upper bound for the squeezing. \textbf{Reconsider these last sentences when we have the figure.  Maybe qualify `lower bound' for constant density models only?}  This means that we can use the Bogoliubov analysis to guide our search for parameter regimes with higher squeezing.

To minimise the damaging effects of the non-zero momentum modes, it is better to operate in a regime where the total occupation of the non-zero momentum modes $N_{\mathbf{k}\neq 0}$ is small, which is also fortuitously the regime in which the Bogoliubov theory is a good approximation.  
The number of excited modes is almost entirely determined by the density of modes energetically accessible by the system.  While this gets larger with the dimensionality and the volume of the system at a given chemical potential, in practice the dimensionality and the volume of the BEC are controlled by changing the trap parameters, which has a dramatic effect on the attainable chemical potential.  So the true figure of merit controlling the degradation of the squeezing due to coupling between the modes is the number of states energetically reachable from the chemical potential.  This scales differently with trap size in each of 1D, 2D and 3D.  For one-dimensional traps, the number of accessible states increases with volume, whereas for 2D traps it is constant with volume. For three-dimensional traps, the number of accessible states \emph{decreases} with volume.

@@

In Fig.~\ref{fig:3DBogScaling} we see that the number of accessible states for three-dimensional traps actually decreases with volume.  This is because the energy of the ground state of the BEC is reducing faster than the density of states at low energies is increasing.  \textbf{Create these figures (by combining old figures) and check both for latest updates.  Confirm that `simulation' (blue line) means TWA, as opposed to Graham's prediction.}  The non-smooth nature of the Bogoliubov calculation is due to the coarse grid, which leads to granularity as the scale factor is changed.
